Splitting of the homology of the punctured mapping class group

### 2019-02-18 Rational.tex
Implemented some corrections by cfb, created gitignore file

\section{A rational counterexample}
In this section we prove that a statement as in theorem \ref{thm:Hbms*as*ggrep}
cannot hold if we consider homology with coefficients in $\Q$.

We recall that $H_*(\cms;\Q)$ has been computed \emph{as a bigraded $\Q-$vector space}
in \cite{BCM} and, more recently, in \cite{Knudson}; a description of it as a
$\gg-$representation seems to be still missing in the literature.

In particular we prove the following fact.
\begin{thm}
 \label{thm:counterexample}
 Let $g\geq 2$ and $m\geq 2$; then $H_2(C_2(\S);\Q)$ is not a symplectic
 representation of $\gg$.
\end{thm}
\begin{proof}
 We will use the following strategy:
 \begin{itemize}
  \item we define a homology class $[a]\in H_2(C_2(\S);\Q)$ represented by a cycle $a$;
  \item we prove that $[a]\neq 0$ by computing the algebraic intersection of
  $[a]$ with a homology class
  in $\tH_2(\cms^c;\Q)$; here it is worth stressing that for all $m$ the manifold
  $\cms$ is orientable: therefore we can apply Poincaré-Lefschetz duality
  also with rational coefficients;
  \item we define another homology class $[b]\in H_2(\cms;\Q)$ and show that
  $[b]$ is mapped to $[b]+2[a]$ by some element in the Torelli group $\mathcal{I}_{g,1}$.
 \end{itemize}
Recall that the Torelli group $\mathcal{I}_{g,1}$ is the kernel of the surjective homomorphism
$\gg\to Sp_{2g}(\Z)$; if an element of the Torelli group acts non-trivially on
some class in $H_2(C_2(\S);\Q)$, then this is not a symplectic representation of $\gg$.

@@

We consider the torus $a=c\times c'$ of configurations in $C_2(\S)$ having one point lying on $c$ and
one lying on $c'$: by abuse of notation, denote also one of its fundamental cycles by $a$
(there are two possible choices, corresponding to the two possible orientations of $a$).

Let $\tup=(0,(u_i,v_i))$ with $u_1=2$ and all other $u_i$'s, as well as all $v_i$'s, equal to zero;
then the map
\[
\phi_{\tup}\colon (\Delta^{\tup},\partial\Delta^{\tup})\to (C_2(\S)^c,\infty)
\]
maps the fundamental class in $H_2(\Delta^{\tup},\partial\Delta^{\tup};\Q)$ to a homology
class in $H_2(C_2(\S);\Q)$; if we see $e^{\tup}$ as a properly embedded, orientable submanifold
of $C_2(\S)$, then we are just considering its fundamental class $[e^{\tup}]\in H_2(C_2(\S)^c)$.

The cell $e^{\tup}$ intersects once, transversely the torus $a$, therefore $[a]\cap[e^{\tup}]=\pm 1$,
where the signs depends on how we have chosen orientations on $a$, $e^{\tup}$ and $C_2(\S)$ itself;
in particular $[a]\neq 0$.

The action of $\gg$ on isotopy classes of simple closed curves is transitive, so we can repeat
the construction of the torus $a$ with any other copy of parallel, non-separating simple closed curves
$c$ and $c'$, and the resulting class $[a]$ will always be non-zero in $H_2(C_2(\S);\Q)$.

Let now $d$ and $d'$ be disjoint, non-separating simple closed curves such that
cutting $\S$ along $d$ and $d'$ we obtain a subsurface $\S'\simeq\Sigma_{1,2}$ of genus 1, with boundary
components $d$ and $d'$ (here we need that the genus of $\S$ is at least 2);
suppose moreover that $c$ is a non-separating simple closed curve in $\S'$,
and let $c',c''$ be two parallel copies of $c$ in $\Sigma_{1,2}$, one on each side of a small tubular
neighborhood of $c$.
Then $d,d',c',c''$ are the boundary of a subsurface $\S''\simeq\Sigma_{0,4}\subset\S'$.
Orient all curves $d,d',c,c',c''$ in such a way that the following equalities hold:
\begin{itemize}
 \item $[d]=[d']\in H_1(\S;\Q)$, as witnessed by the subsurface $\S'$;
 \item $[c]=[c']=[c'']\in H_1(\S;\Q)$;
 \item $[d]-[d']+[c']-[c'']=0\in H_1(\S\setminus c;\Q)$, as witnessed by the subsurface $\S''$.
\end{itemize}
See picture PICTURE.

@@

Let again $a=[c\times c']$ and let $b=[c\times d]$: then the class $b'=[c\times d']$
is equal to $b+2a\in H_2(C_2(\S);\Q)$.

Consider now an element of the Torelli group that fixes $c$ and maps $d$ to $d'$ preserving
the orientation, for example the bounding pair represented in picture PICTURE:
then the class $b$ is mapped to $b'=b+2a\neq b$.
\end{proof}

>>>>

\section{A rational counterexample}
\label{sec:rational}
In this section we prove that a statement as in Theorem \ref{thm:Hbms*as*ggrep}
cannot hold if we consider homology with coefficients in $\Q$. We do not know
if the analogue of Theorem \ref{thm:main} holds in homology with coefficients in $\Q$
or in fields of odd characteristic.

We point out that $H_*(C_m(\Sigma_{g,1});\Q)$ has been computed \emph{as a bigraded $\Q$-vector space}
by B\"odigheimer, Cohen and Milgram in \cite{BCM}, and more recently by Knudson in \cite{Knudson}. A description of
these homology groups as a
$\gg$-representation seems to be still missing in the literature.

We will prove the following theorem:
\begin{thm}
 \label{thm:counterexample}
 Let $g\geq 2$ and $m\geq 2$; then $H_2(C_2(\Sigma_{g,1});\Q)$ is not a symplectic
 representation of $\gg$.
\end{thm}
\begin{proof}
 Let again $\S=\Sigma_{g,1}$. We will use the following strategy:
 \begin{itemize}
  \item we define a homology class $[a]\in H_2(C_2(\S);\Q)$ represented by a cycle $a$;
  \item we prove that $[a]\neq 0$ by computing the algebraic intersection of
  $[a]$ with a homology class
  in $\tH_2(\cms^{\infty};\Q)$: note that the manifold
  $\cms$ is orientable, hence Poincaré-Lefschetz duality holds
  also with rational coefficients;
  \item we define another homology class $[b]\in H_2(\cms;\Q)$ and show that
  $[b]$ is mapped to $[b]+2[a]$ by some element in the Torelli group $\mathcal{I}_{g,1}$.
 \end{itemize}
Recall that the Torelli group $\mathcal{I}_{g,1}$ is the kernel of the surjective homomorphism
$\gg\to Sp_{2g}(\Z)$ induced by the action of $\gg$ on $H_1\pa{\sg;\Z}$; if an element of the Torelli group acts non-trivially on
some class in $H_2(C_2(\S);\Q)$, the latter cannot be a symplectic representation of $\gg$.

@@

We consider the torus $a=c\times c'$ of configurations in $C_2(\S)$ having one point lying on $c$ and
one lying on $c'$; we let $[a]\in H_1(C_2(\S);\Q)$ be the fundamental class of $a$ associated with one
orientation of $a$.

Let $\tup=(0,\uu,\uv)$ with $\uu=(2,0)$ and $\uv=(0,0)$.
Then the map
\[
\Phi_{\tup}\colon (\Delta^{\tup},\partial\Delta^{\tup})\to (C_2(\S)^{\infty},\infty)
\]
maps the fundamental class in $H_2(\Delta^{\tup},\partial\Delta^{\tup};\Q)$ to a
class in $\tH_2(C_2(\S)^{\infty};\Q)$, that we call $[\kappa(0,\uu,\uv)]_{\Q}$.
% ; if we see $e^{\tup}$ as a properly embedded, orientable submanifold
% of $C_2(\S)$, then we are just considering its fundamental class $[e^{\tup}]\in H_2(C_2(\S)^{\infty})$.

The cell $e^{\tup}$ intersects once, transversely the torus $a$, therefore the
algebraic intersection $[a]\cap[\kappa(0,\uu,\uv)]_{\Q}$ is $\pm 1$,
where the signs depends on how we have chosen orientations on $a$, $e^{\tup}$ and $C_2(\S)$ itself.
In particular $[a]\neq 0$.

The action of $\gg$ on isotopy classes of non-separating simple closed curves is transitive, so we can repeat
the construction of the torus $a$ with any other copy of parallel, non-separating simple closed curves
$c$ and $c'$, and the resulting class $[a]$ will always be non-trivial in $H_2(C_2(\S);\Q)$.

See Figure \ref{fig:rational} to visualize the following discussion.
Let $d$ and $d'$ be disjoint, non-separating simple closed curves such that
the following holds: if we cut $\S$ along $d$ and $d'$ we separate $\S$ into two pieces, one of
which is a subsurface $\S'\simeq\Sigma_{1,2}$ of genus 1, with boundary
components $d$ and $d'$. Here we use that the genus of $\S$ is at least 2.

Suppose now that $c$ is a non-separating simple closed curve in $\S'$,
and let $c',c''$ be two parallel copies of $c$ in $\Sigma_{1,2}$, one on each side of a small tubular
neighborhood of $c$.
Then $d,d',c',c''$ are the boundary of a subsurface $\S''\simeq\Sigma_{0,4}\subset\S'$.
Orient all curves $d,d',c,c',c''$ in such a way that the following equalities hold:
\begin{itemize}
 \item $[d]=[d']\in H_1(\S;\Q)$, as witnessed by the subsurface $\S'$;
 \item $[c]=[c']=[c'']\in H_1(\S;\Q)$;
 \item $[d]-[d']+[c']-[c'']=0\in H_1(\S\setminus c;\Q)$, as witnessed by the subsurface $\S''$.
\end{itemize}

\begin{figure}\centering
 \includegraphics[scale=1.0]{figures/rational.pdf}
 \caption{The curves $d,d',c,c',c'',\gamma,\gamma''$ and the subsurfaces $\S',\S''$.}
\label{fig:rational}
\end{figure}

@@

Let again $[a]=[c\times c']$ and let $[b]=[c\times d]$: then the class $[b']=[c\times d']$
is equal to $[b]+2[a]\in H_2(C_2(\S);\Q)$.

Consider now an element of the Torelli group that fixes $c$ and maps $d$ to $d'$ preserving
the orientation, for example the bounding pair $D_{\gamma}\circ D_{\gamma'}^{-1}$, where
$\gamma$ and $\gamma'$ are represented in Figure \ref{fig:rational}:
then the class $b$ is mapped to $b'=b+2a\neq b$.
\end{proof}

The previous proof works verbatim if we replace $\Q$ by any field of odd characteristic. Moreover
all the arguments in the previous proof can be adapted to show that $H_*(F_2(\S);\mathbb{F})$ is
not a symplectic representation of $\gg$ (see Definition \ref{defn:cms}), where
$\mathbb{F}$ is \emph{any} field, including $\Z_2$.

Therefore, at least for orientable surfaces $\sg$ of genus $g\geq 2$,
being a symplectic representation of the mapping class group seems to be a peculiarity
of the homology of \emph{unordered} configuration space and of the characteristic 2.

### 2019-02-18 Ssequence.tex
Implemented some corrections by cfb, created gitignore file

\section{Proof of theorem \ref{thm:main}}
We will prove theorem \ref{thm:main} by induction on $m$.
The case $m=0$ is trivial, so suppose $m\geq 1$.

For all $m\geq 0$ we let $E(m)$ be the Leray-Serre spectral sequence associated with
the bundle \ref{eq:BirmanbundleD}: its second page has the form
\[
 E(m)^2_{k,q}=H_k(B\Diff(\S;\partial\S\cup\D);H_q(\cms)).
\]
From theorem \ref{thm:Hbms*as*ggrep} we know that this spectral sequence is concentrated
on the rows $q=0,\dots, m$.

We want to prove the vanishing of all differentials appearing in the pages
$E(m)^r$ with $r\geq 2$; the $r-$th differential takes the form
\[
 \partial_r\colon E(m)^r_{k,q}\to E(m)^r_{k-r,q+r-1}.
\]
In particular any differential $\partial_r$ exiting from the row $q=m$ is trivial,
because it lands in a higher, hence trivial row.

Fix now $q=m-l<m$, in particular $l\geq 1$; by theorem \ref{thm:Hbms*as*ggrep}, and
in particular by lemma \ref{lem:oplussplitting}, we have a splitting of $H_k(\gg;H_{m-l}(\cms))$ as
\begin{equation}
\label{eq:Hksplitting}
 \bigoplus_{p=l}^m H_k\pa{\gg;\mu_*\pa{H_{p-l}(C_p(\D))\otimes H_{m-p}(C_{m-p}(\S'))}}.
\end{equation} 
We fix now $l\leq p\leq m$ and show the vanishing of all differentials $\partial_r$ exiting from the
summand with label $p$ in the previous equation.

Consider the map $\mu^{\cF}$ from definition \ref{defn:universalSbundle} as a map of
bundles over the space $B\Diff(\S;\partial\S\cup\D)$:
\[
 \mu^{\cF}\colon C_p(\D)\times C_{m-p}(\cF_{\S'})\to C_m(\cF_{\S,\D}).
\]
Notice that the first bundle $C_p(\D)\times C_{m-p}(\cF_{\S'})\to B\Diff(\S;\partial\S\cup\D)$
is the product of the space $C_p(\D)$ with the bundle $C_{m-p}(\cF_{\S'})\to B\Diff(\S;\partial\S\cup\D)$;
therefore the spectral sequence associated with $C_p(\D)\times C_{m-p}(\cF_{\S'})$ is isomorphic,
from the second page on, to the tensor product
of $H_*(C_p(\D))$ and the spectral sequence associated with the bundle $C_{m-p}(\cF_{\S'})$; the
latter spectral sequence is isomorphic, in our notation, to the spectral sequence $E(m-p)$.
In particular $\mu^{\cF}$ induces a map of spectral
sequences
\[
\mu^{\cF}_*\colon H_*(C_p(\D))\otimes E(m-p)\to E(m);
\]
that in the second page, on the $(m-l)-$th row, restricts to
the inclusion of one of the direct summands in equation \ref{eq:Hksplitting}:
\[
H_k\pa{\gg;\mu_*(H_{p-l}(C_p(\D))\otimes H_{m-p}(C_{m-p}(\S')))}\subset H_k\pa{\gg;H_{m-l}(\cms)}.
\]

@@

One can generalise theorem \ref{thm:main} to orientable or non-orientable
surfaces with at least one boundary component, following the generalisation
of theorem \ref{thm:Hbms*as*ggrep} discussed at the end of section \ref{sec:Actiongg}.

>>>>

\section{Proof of Theorem \ref{thm:main}}
We will prove Theorem \ref{thm:main} by induction on $m$.
The case $m=0$ is trivial.

For all $m\geq 0$ we let $E(m)$ be the Leray-Serre spectral sequence associated with
the bundle \eqref{eq:BirmanbundleD}: its second page has the form
\[
 E(m)^2_{k,q}=H_k(B\Diff(\S;\partial\S\cup\D);H_q(\cms))=H_k(\gg;H_q(\cms)).
\]
From Theorem \ref{thm:Hbms*as*ggrep} we know that this spectral sequence is concentrated
on the rows $q=0,\dots, m$.

We want to prove the vanishing of all differentials appearing in the pages
$E(m)^r$ with $r\geq 2$; the $r$-th differential takes the form
\[
 \partial_r\colon E(m)^r_{k,q}\to E(m)^r_{k-r,q+r-1}.
\]
In particular any differential $\partial_r$ exiting from the row $q=m$ is trivial,
because it lands in a higher, hence trivial row.

Fix now $q=m-l<m$, in particular $l\geq 1$; by Theorem \ref{thm:Hbms*as*ggrep}, and
in particular by Lemma \ref{lem:oplussplitting}, we have a splitting of $H_k(\gg;H_{m-l}(\cms))$ as
\begin{equation}
\label{eq:Hksplitting}
 \bigoplus_{p=l}^m H_k\pa{\gg;\mu_*\pa{H_{p-l}(C_p(\D))\otimes H_{m-p}(C_{m-p}(\S'))}}.
\end{equation} 
We fix now $l\leq p\leq m$ and show the vanishing of all differentials $\partial_r$ exiting from the
summand with label $p$ in the previous equation.

Consider the map $\mu^{\cF}$ from Definition \ref{defn:universalSbundle} as a map of
bundles over the space $B\Diff(\S;\partial\S\cup\D)$:
\[
 \mu^{\cF}\colon C_p(\D)\times C_{m-p}(\cF_{\S'})\to C_m(\cF_{\S,\D}).
\]
Note that the first bundle $C_p(\D)\times C_{m-p}(\cF_{\S'})\to B\Diff(\S;\partial\S\cup\D)$
is the product of the space $C_p(\D)$ with the bundle $C_{m-p}(\cF_{\S'})\to B\Diff(\S;\partial\S\cup\D)$;
therefore the spectral sequence associated with $C_p(\D)\times C_{m-p}(\cF_{\S'})$ is isomorphic,
from the second page on, to the tensor product
of $H_*(C_p(\D))$ and the spectral sequence associated with the bundle $C_{m-p}(\cF_{\S'})$; the
latter spectral sequence is isomorphic, in our notation, to the spectral sequence $E(m-p)$.
In particular $\mu^{\cF}$ induces a map of spectral
sequences
\[
\mu^{\cF}_*\colon H_*(C_p(\D))\otimes E(m-p)\to E(m);
\]
that in the second page, on the $(m-l)$-th row and $k$-th column, restricts to
the inclusion of one of the direct summands in equation \eqref{eq:Hksplitting}:
\[
H_k\pa{\gg;\mu_*(H_{p-l}(C_p(\D))\otimes H_{m-p}(C_{m-p}(\S')))}\subset H_k\pa{\gg;H_{m-l}(\cms)}.
\]

@@

One can generalise Theorem \ref{thm:main} to orientable or non-orientable
surfaces with non-empty boundary, following the generalisation
of Theorem \ref{thm:Hbms*as*ggrep} discussed at the end of section \ref{sec:Actiongg}.
